---
title: "Microplate data analysis"
author: "Guilherme Marcelino Viana de Siqueira and María-Eugenia Guazzaroni"
date: "November, 2022"
output:
  html_document:
    css: css/style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this analysis

This analysis depends on the `data/growth_curves` directory found in the root directory for this project and on the `mipreadr` *ad-hoc* package that is available through [GitHub](github.com/viana-guilherme/mipreadr). In each experiment, the plate contained at least two biological replicates (from different colonies) of *E. coli* DH10B and/or *P. putida* KT2440 and three technical replicates (repeated Wells) for each biological replicate. The `.txt` files in the data directory come directly from a Perkin Elmer Victor X3 machine set to read absorbance and fluorescence at periodic time points. The `.xlsx` files are the plate designs used in each experiment, and are needed for the mipreadr workflow

## Setting up the analysis

To start the analysis, we must first load the required packages. If needed, uncomment the line starting with `remotes::install_github()` line to install the development version of mipreadr before loading it.

```{r results='hide', error=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)    # loads the tidyverse...

#remotes::install_github("viana-guilherme/mipreadr")
library(mipreadr)     # loads the analysis package

# the packages below are also through this analysis, be sure to have those installed
library(ggtext)       # used for plotting
library(gt)           # used for helping define the color palettes
library(ggh4x)        # used for plotting
library(multcompView) # used for getting significance levels as letters in the Tukey test 
```

## Importing data

This is straightforward using the mipreadr workflow:

```{r message=FALSE}

# reading the VictorX3 data into R using the mipreadr function addPlate()
exp01 <- addPlate(file = "../data/growth_curves/2212.txt",
                  layout = "../data/growth_curves/2212.xlsx",
                  mode = "x3",
                  normalizeFluorescence = TRUE)

exp02 <- addPlate(file = "../data/growth_curves/2213.txt",
                  layout = "../data/growth_curves/2213.xlsx",
                  mode = "x3",
                  normalizeFluorescence = TRUE)

exp03 <- addPlate(file = "../data/growth_curves/2220.txt",
                  layout = "../data/growth_curves/2220.xlsx",
                  mode = "x3",
                  normalizeFluorescence = TRUE)

# preparing the list of experiments for analysis
exp_list <- list(exp01, exp02, exp03)

# collapsing the eperiment data with mergePlates()
analysis <- mergePlates(exp_list)

# checking the first few lines of the resulting data frame
head(analysis)
```

## 2. Visualizing growth and normalized fluorescence curves

Once we have successfully read and pre-processed the data, we can explore the results using a variety of visualizations. In the code blocks below, we will plot the growth and the normalized fluorescence curves for the entire experiment.

To start things off, we will append some metadata - including a custom color palette - onto the analysis data frame so that we have more flexibility for plotting:

```{r}

### Define a manual color vector for coloring the plots consistently ----
basecolors <- RColorBrewer::brewer.pal(9, name = "Spectral") |>
  gt::adjust_luminance(steps = -0.1)

colorPalette <- basecolors[c(3:1, 7:9)]

# the modifications below are mainly cosmetic

analysis <- analysis |>
  #this part helps making the names more tidy for the plot
  mutate(uniqueID = glue::glue("{Sample}:{Condition}"),
        readableName = case_when(
              Sample == "pSEVA" ~ "pSEVA231-∅",
              Sample == "pSEVA.GFP" ~ "pSEVA231-*gfplva*",
              Sample == "pSEVA.P100.GFP" ~ "pSEVA231-P*j100*-*gfplva*",
              Sample == "pVANT" ~ "pVANT-∅",
              Sample == "pVANT.GFP" ~ "pVANT-*gfplva*",
              Sample == "pVANT.P100.GFP" ~ "pVANT-P*j100*-*gfplva*"
              ) %>%
          # this part gives us a way to reorder the columns in the barplot
          as_factor() %>%
          fct_relevel(c("pSEVA231-∅",
                        "pSEVA231-*gfplva*",
                        "pSEVA231-P*j100*-*gfplva*",
                        "pVANT-∅",
                        "pVANT-*gfplva*",
                        "pVANT-P*j100*-*gfplva*")
          ),
        # this part helps us plot the species names in italics using ggtext
        readableCondition = case_when(
              Condition == "Ecoli" ~ "*E. coli* DH10B",
              Condition == "Pputida" ~ "*P. putida* KT2440",
              ),
        # this part attaches a specific color from our palette to a given sample
        light_color = case_when(
          readableName == "pSEVA231-∅" ~ colorPalette[1],
          readableName == "pSEVA231-*gfplva*" ~ colorPalette[2],
          readableName == "pSEVA231-P*j100*-*gfplva*" ~ colorPalette[3],
          readableName == "pVANT-∅" ~ colorPalette[4],
          readableName == "pVANT-*gfplva*" ~ colorPalette[5],
          readableName == "pVANT-P*j100*-*gfplva*" ~ colorPalette[6]
          ),
        # this part provides a darker version of the color palette for specific effects
        dark_color = adjust_luminance(light_color, -1)
  )
```

We also will define named `color` and `fill` vectors to consistently color our samples in the plots based on the dark_color and light_color columns we just created

```{r}

color_vector <- analysis |>
  select(readableName, dark_color) |>
  distinct() |>
  deframe()

fill_vector <- analysis |>
  select(readableName, light_color) |>
  distinct() |>
  deframe()

```

Finally, we can use `ggplot2` to plot the curves. First OD600 *versus* time:

```{r fig.width=10, fig.align='center'}

analysis |>
  ggplot(aes(x = Repeat, y = `Absorbance@600_Mean`)) +
  geom_point(aes(color = readableName, fill = readableName),
                 stroke = 0.7, size = 3, alpha = 0.8) +
  geom_ribbon(aes(ymin= `Absorbance@600_Mean` - `Absorbance@600_SD`,
                  ymax= `Absorbance@600_Mean` + `Absorbance@600_SD`,
                  fill = readableName),
              alpha = 0.1) +
  
  # using a semilogarithmic scale with custom breaks
  scale_y_continuous(trans = "log", breaks = c(0.005, 0.02, 0.08, 0.32)) +
  # the line below converts the timepoints (x-axis) into half-hour periods
  scale_x_continuous(labels = ~ {(.x - 1)/2}, breaks = seq(from = 1, to = 17, by = 4)) +
  
  # Editing some of the visual aspects of the plot
  labs(x = "Time (hours)", y = "OD<sub>600</sub>") +
  guides(fill = guide_legend(title = "", ncol = 1, byrow = TRUE),
         color = guide_legend(title = "", ncol = 1, byrow = TRUE),
         shape = guide_legend(title = "", ncol = 1, byrow = TRUE),) +
  # below we use our custom palette vectors. In this instance we want color and fill to be the same,
  # so we will repeat it...
  scale_color_manual(values = fill_vector) +
  scale_fill_manual(values = fill_vector) +
  
  # separates the plot between species
  facet_wrap(~ readableCondition,  strip.position = "top", nrow = 2) +
  
  # setting up the theme
  theme_light(base_size = 16) +
  theme(
        axis.title.x = element_markdown(size = rel(1.3), color = "black"),
        axis.text.x = element_markdown(size = rel(1.2), color = "black"),
        axis.title.y = element_markdown(size = rel(1.3), color = "black"),
        axis.text.y = element_markdown(size = rel(1.2), color = "black"),

        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.margin = unit(c(30,10,10,20), "pt"),

        legend.spacing.y = unit(10, "pt"),
        legend.spacing.x = unit(15, "pt"),
        legend.box.spacing = unit(0, "pt"),
        legend.position = "right",
        legend.key = element_rect(color = "grey60", size = 0.1),
        legend.key.size = unit(25, "pt"),
        legend.box.margin = margin(12,0,0,0, unit = "pt"),
        legend.text = ggtext::element_markdown(size = rel(1.05)),

        strip.background = element_blank(),
        strip.text = ggtext::element_markdown(color = "black", size = rel(1.1)),
        strip.placement = "outside"
        )

```

We can similarly plot the normalized fluorescence curve:

```{r fig.width=10, fig.align='center'}
analysis |>
  ggplot(aes(x = Repeat, y = `norm.GFP_Mean`)) +
  geom_point(aes(color = readableName, fill = readableName),
             stroke = 0.7, size = 4, alpha = 0.8) +
  geom_ribbon(aes(ymin = `norm.GFP_Mean` - `norm.GFP_SD`,
                  ymax = `norm.GFP_Mean` + `norm.GFP_SD`,
                  fill = readableName),
              alpha = 0.1) +
  scale_y_continuous(trans = "log10") +
  scale_x_continuous(labels = ~ {(.x - 1)/2}, breaks = seq(from = 1, to = 17, by = 4)) +
  labs(x = "Time (hours)", y = "Normalized Fluorescence (log<sub>10</sub>)") +
  guides(fill = guide_legend(title = "", ncol = 1, byrow = TRUE),
         color = guide_legend(title = "", ncol = 1, byrow = TRUE),
         shape = guide_legend(title = "", ncol = 1, byrow = TRUE),) +
  scale_color_manual(values = fill_vector) +
  scale_fill_manual(values = fill_vector) +
  facet_wrap(~ readableCondition,  strip.position = "top", nrow = 2) +
  theme_light(base_size = 16) +
  theme(axis.title.x = ggtext::element_markdown(size = rel(1.3), color = "black"),
        axis.text.x = ggtext::element_markdown(size = rel(1.2), color = "black"),
        axis.title.y = ggtext::element_markdown(size = rel(1.3), color = "black"),
        axis.text.y = ggtext::element_markdown(size = rel(1.2), color = "black"),

        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.margin = unit(c(30,10,10,20), "pt"),

        legend.spacing.y = unit(10, "pt"),
        legend.spacing.x = unit(15, "pt"),
        legend.box.spacing = unit(0, "pt"),
        legend.position = "right",
        legend.key = element_rect(color = "grey60", size = 0.1),
        legend.key.size = unit(25, "pt"),
        legend.box.margin = margin(12,0,0,0, unit = "pt"),
        legend.text = ggtext::element_markdown(size = rel(1.05)),

        strip.background = element_blank(),
        strip.text = ggtext::element_markdown(color = "black", size = rel(1.1)),
        strip.placement = "outside")
```

## Investigating the fluorescence at 4h

To gain a better understanding of the differences between our bacterial strains carrying different plasmids, we will look in more detail at an intermediate point in the growth curve. The "Repeat" column in the dataset refers to the point collected by the machine in different moments of the experiment (1 being the first timepoint at To, 2 being the point taken after 30 minutes, 3 being the point taken after one hour...). We can choose the timepoint at 4 hours and save it in a `choosePoint` variable, so that we have a convenient way to subset our data.

```{r}
# selecting the Repeat value that corresponds to the fourth hour of the experiment
choosePoint <- 5

```

Now we may proceed to the statistical analysis of the data. For this, after subsetting the original data to only include the point where Repeat is equal to choosePoint, we will adjust a linear model and perform the Tukey post-hoc test for multiple comparisons. With this, we will create a `tukey.lcd` with the letter codes for each significance group in our analysis.

```{r}

# collect the individual replicates into a single table
anova_dataset <- exp_list %>%
  purrr::map_dfr(4) %>% # return the fourth table within every element of the exp_list (the results)
  dplyr::filter(Repeat == choosePoint) %>% # we are interested only the fifth repeat in the experiments
  dplyr::select(Sample, Condition, norm.GFP)

# adjust the linear model
anova <- aov(formula = log(norm.GFP) ~ Sample * Condition, data = anova_dataset)

# perform post-hoc correction and pairwise comparisons
tukey <- TukeyHSD(anova)

# convert the p-values to letter codes
tukey.lcd <- multcompView::multcompLetters4(object = anova, comp = tukey, reversed = TRUE) %>%
  purrr::pluck(3) %>% # extracting the information from the
  purrr::pluck(1) %>% # deeply nested resulting file structure
  enframe(name = "uniqueID", value = "tukeyCode") # save as a 2-column tibble

# visualizing summary information for the ANOVA model
summary(anova)
```

After obtaining the ANOVA results, we may visualize the resulting information alongside the means for normalized fluorescence as below:

```{r fig.width=7, fig.align='center', warning=FALSE, message=FALSE}

# first, let's define a "factor" with which to scale our data
factor <- 1e4 # the factor allow us to make the numbers more manageable in the visualization
factor_label <- str_count(factor, '0') # this will help us display the multiplication factor in a tidy way in the plot

# now, onto the plotting:

# we start by preprocessing the data a little
analysis |>
  # once again, we are only interested in the Repeat of interest
  dplyr::filter(Repeat == choosePoint) |>
  # we may simplify the values by dividing them by the "factor" set above
  mutate(norm.GFP_Mean = norm.GFP_Mean/factor,
         norm.GFP_SD = norm.GFP_SD/factor) |>
  # add the tukey tet results to the dataset for convenient plotting
  left_join(tukey.lcd) |>
  
  # now we may call ggplot, as usual
  ggplot(aes(
    x = readableName,
    y = norm.GFP_Mean,
    fill = readableName)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  #add the standard deviation bars
  geom_errorbar(aes(ymin = norm.GFP_Mean - norm.GFP_SD,
                    ymax = norm.GFP_Mean + norm.GFP_SD),
                stat = "identity",
                position = "dodge",
                width = 0.1,
                color = "black") +
  # cosmetic changes to the plot
  guides(fill = guide_legend(title = "", byrow = TRUE), x = "axis_truncated") +
  scale_color_manual(values = color_vector) +
  scale_fill_manual(values = fill_vector) +
  facet_wrap(~ readableCondition,  strip.position = "bottom") +
  scale_y_continuous(n.breaks = 6) +
  # adding the custom y-axis title
  labs(x = "", y = glue::glue("Normalized fluorescence (x10^{factor_label} A.U.)")) +
  theme_light(base_size = 20) +
  theme(axis.text.x = element_blank(),
        axis.title.y = ggtext::element_markdown(size = rel(0.8)),
        axis.text.y = ggtext::element_markdown(size = rel(1), color = "black"),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.spacing.y = unit(2, "pt"),
        legend.box.spacing = unit(0, "pt"),
        legend.position = c(0.265,.85),
        plot.margin = unit(c(20,0,-10,20), "pt"),
        axis.line.x.bottom = element_line(color = 'black'),
        axis.line.y.left = element_line(color = 'black', size = 0.8),
        panel.border = element_blank(),
        panel.spacing.x = unit(20, "pt"),
        legend.text = ggtext::element_markdown(),
        strip.background = element_blank(),
        strip.text = ggtext::element_markdown(color = "black"),
        strip.placement = "outside")


```
